The large scale clustering in dispersion measure space

### 2015-05-20 writeup/dm_space_distortions.tex
Filled out some text.

% This paragraph needs to be reorganized.
However, redshift is not the only measure of radial distance. In particular,
dispersion of electromagnetic signals due to intervening plasma
is readily measured from fast radio bursts (FRB) which are a class
of radio transient first reported in \citet{2007Sci...318..777L}.
There have since been 
10 such bursts detected \citep{2012MNRAS.425L..71K,2013Sci...341...53T, 2014ApJ...790..101S, 2014ApJ...792...19B, 2015MNRAS.447..246P,2015ApJ...799L...5R} and total rate of bursts
been estimated to be $3.3^{+5}_{-2.5} \times 10^3$
per sky per day \cite{2015arXiv150500834R}. 
The dispersion measure
(DM) is a line-of-site integral of the free electron density and can thus be
used as a distance estimate.
If the progenitors of these bursts are at cosmological distances, with the
majority of the dispersion caused by electrons in the inter-galactic medium,
then the average DM-distance relation can be modelled
\citep{2004MNRAS.348..999I, 2003ApJ...598L..79I}.

@@

Large-scale structure is usually studied through it two-point statistics, most
commonly the power-spectrum, $P(k)$.
Unlike in redshift-space distortions, Equation~\ref{e:delta_s} does not have a
simple form in harmonic space. The equation's third term couples harmonic modes
and thus the two-point statistics cannot be phrased as a simple power spectrum.
We will instead use $C^{ss}_\ell(\chi,\chi')$, which is
the cross-correlation angular power-spectrum of the DM space over-density
on shells at $\chi$ and $\chi'$.
\begin{align}
    &\delta_{\ell\ell'}\delta_{mm'}C^{ss}_\ell(\chi, \chi') = \nonumber \\
    & \quad
        \left\langle
        \int\ud\Omega Y_{\ell m}(\hat n)
       \delta_{fs}(\hat n \chi, \chi)
        \int\ud\Omega' Y_{\ell' m'}(\hat n')
        \delta_{fs}(\hat n' \chi', \chi')
        \right\rangle
\end{align}
The first two term in Equation~\ref{e:delta_s} are stationary. For stationary,
isotropic, tracers $x$ and $y$ we have 
$\langle \delta_x(\vec k, \chi) \delta_y(\vec k', \chi) \rangle = (2\pi)^3
\delta^3(\vec k - \vec k') P_{xy}(k, \chi)$.  If for a moment we ignore time
dependance, the angular cross power spectrum of such tracers is
\begin{align}
    %\delta_{\ell\ell'}\delta_{mm'}C^{xy}_\ell(\chi, \chi')
    %&= \int\ud\Omega\ud\Omega'Y_{\ell m}(\hat n) Y_{\ell' m'}(\hat n')
    %    \langle \delta_x(\hat n \chi) \delta_y(\hat n' \chi') \rangle,
    %    \\
    C^{xy}_\ell(\chi,\chi')
    &= \frac{2}{\pi}
\int_0^\infty\ud k k^2 j_\ell(k\chi) j_{\ell}(k\chi')P_{xy}(k).
\end{align}
In reality the power spectrum evolves on the order of a Hubble time.
The angular cross-correlations will be very small unless $\chi$ and $\chi'$ are
within a few correlation lengths of one another, roughly a few hundred
megaparsecs.  The evolution of the power-spectrum is negligible over these time
differences which leads to a straight forward way to include the evolution of
the perturbations:
\begin{align}
C^{xy}_\ell(\chi,\chi') 
    \approx \frac{2}{\pi}
    \int_0^\infty\ud k k^2
    j_\ell(k\chi) j_{\ell}(k\chi')
    P_{xy}(k,(\chi + \chi')/2),\nonumber &\\
    |\chi - \chi'| \ll 1/aH.&
\end{align}

@@

Equation~\ref{e:Clss} can be simplified substantially by adopting the small
angle and Limber approximations \citep{1953ApJ...117..134L,
1992ApJ...388..272K, 1998ApJ...498...26K}.  The small angle
approximation eliminates the $k$ integral over spherical Bessel functions, replacing it
with a Fourier transform, and is valid for $\ell \gg 1$. The Limber
approximation assumes that only modes with small radial component of their wave
vector contribute to the radial integrals and is valid if the power spectra
evolve slowly compared to the correlation length \citep{2008PhRvD..78l3506L}
(which has already been
assumed). With these approximations we have
\begin{align}
    &C^{ss}_\ell(\chi, \chi') \approx
    \nonumber \\ & \quad
    \frac{1}{\bar\chi^2}
        \int_{-\infty}^\infty\frac{\ud k_\parallel}{(2 \pi)} 
        e^{i k_\parallel (\chi - \chi')}
        P_{[ee + ff - 2ef]}
        (\sqrt{k_\parallel^2 + \nu^2/\bar\chi^2}, \bar\chi)
    \nonumber \\ & \quad +
    A(\chi) A(\chi')
    \int_0^{\chi_{\rm min}}\ud\chi''
    \frac{1}{\chi^{\prime\prime 2}}
    P_{ee}(\nu/\chi'',\chi'')
    \nonumber \\ & \quad +
    \frac{A(\chi_{\max})}{\chi_{\min}^2}
    P_{[ee - ef]}(\nu/\chi_{\min}, \chi_{\min}),
    \label{e:Clss_limber}
\end{align}
where $\nu\equiv\ell + 1/2$, $\chi_{\min} \equiv \min(\chi, \chi')$,
and $\chi_{\max} \equiv \max(\chi,
\chi')$ .  We've found that these approximations are excellent for $\ell
\gtrsim 10$
and use this form for the remainder of the paper.

@@

Description of the input power-spectra, $\bar n_f$ and figures.
Figure~\ref{f:n_f}, Figure~\ref{f:terms}.

\begin{figure}
    \includegraphics[scale=0.42]{figures/n_f}
    \caption{
        \label{f:n_f}
        \emph{Top:} Comoving density of sources for toy model, as given in
        Equation~\ref{e:n_f}. Normalization is
        such that the survey has a total of $10^4$ FRBs.
        \emph{Bottom:} Coefficient $A(\chi)$ (defined in Equation~\ref{e:A})
        in toy model.
    }
\end{figure}

\begin{figure}
    \includegraphics[scale=0.42]{figures/terms}
    \caption{
        \label{f:terms}
        Terms in the DM-space cross-correlation angular power spectrum.
        All terms are evaluated at $\bar\chi=1000\,{\rm MPc}/h$ and 
        $\chi - \chi' = 10\,{\rm MPc}/h$, except the lowest most 
        ``local term'' curve which is for $\chi - \chi'=50\,{\rm MPc}/h$. 
        While
        the other two terms are highly insensitive to the separation of
        the radial slices, the local term drops rapidly at high separations.
    }
\end{figure}

@@

>>>>

% This paragraph needs to be reorganized.
However, redshift is not the only measure of radial distance. In particular,
dispersion of electromagnetic signals due to intervening plasma
is readily measured from fast radio bursts (FRB) which are a class
of radio transient first reported in \citet{2007Sci...318..777L}.
There have since been 
10 such bursts reported \citep{2012MNRAS.425L..71K,2013Sci...341...53T, 2014ApJ...790..101S, 2014ApJ...792...19B, 2015MNRAS.447..246P,2015ApJ...799L...5R} and total rate of bursts
been estimated to be $3.3^{+5}_{-2.5} \times 10^3$
per sky per day \cite{2015arXiv150500834R}. 
The dispersion measure
(DM) is a line-of-site integral of the free electron density and can thus be
used as a distance estimate, as is routinely done for galactic pulsars.
Because FRBs are observed to have dispersion measures greatly in excess of
the galactic expectation they must be of extra-galactic origin, with their
dispersion produced either by
inter-galactic medium (IGM) \citep{2013Sci...341...53T} or the
environment of the source
\citep{2014ApJ...785L..26L,2015arXiv150101341P}.
If the dispersion is dominated by the inter-galactic medium, then 
 then the average DM-distance relation can be modelled
\citep{2004MNRAS.348..999I, 2003ApJ...598L..79I} and the sources are at
cosmological distances of a few gigaparsecs. This would permit
the study of the clustering of FRBs in three dimensions which should trace the
large-scale structure of the Universe on linear scales.

@@

Large-scale structure is usually studied through its two-point statistics, most
commonly the power-spectrum, $P(k)$.
Unlike in redshift-space distortions, Equation~\ref{e:delta_s} does not have a
simple form in harmonic space. The equation's third term couples harmonic modes
and thus the two-point statistics cannot be phrased as a simple power spectrum.
We will instead use $C^{ss}_\ell(\chi,\chi')$, which is
the cross-correlation angular power-spectrum of the DM space over-density,
on shells at $\chi$ and $\chi'$.
\begin{align}
    &\delta_{\ell\ell'}\delta_{mm'}C^{ss}_\ell(\chi, \chi') = \nonumber \\
    & \quad
        \left\langle
        \int\ud\Omega Y_{\ell m}(\hat n)
       \delta_{fs}(\hat n \chi, \chi)
        \int\ud\Omega' Y_{\ell' m'}(\hat n')
        \delta_{fs}(\hat n' \chi', \chi')
        \right\rangle
\end{align}
The first two term in Equation~\ref{e:delta_s} are stationary. For stationary,
isotropic, tracers $x$ and $y$ we have 
$\langle \delta_x(\vec k, \chi) \delta_y(\vec k', \chi) \rangle = (2\pi)^3
\delta^3(\vec k - \vec k') P_{xy}(k, \chi)$.  If for a moment we ignore time
dependance, the angular cross power spectrum of such tracers is
\begin{align}
    %\delta_{\ell\ell'}\delta_{mm'}C^{xy}_\ell(\chi, \chi')
    %&= \int\ud\Omega\ud\Omega'Y_{\ell m}(\hat n) Y_{\ell' m'}(\hat n')
    %    \langle \delta_x(\hat n \chi) \delta_y(\hat n' \chi') \rangle,
    %    \\
    C^{xy}_\ell(\chi,\chi')
    &= \frac{2}{\pi}
\int_0^\infty\ud k k^2 j_\ell(k\chi) j_{\ell}(k\chi')P_{xy}(k).
\end{align}
In reality the power spectrum evolves on the order of a Hubble time.
The angular cross-correlations will be very small unless $\chi$ and $\chi'$ are
within a few correlation lengths of one another, roughly a few hundred
megaparsecs.  The evolution of the power-spectrum is negligible over these time
differences which leads to a straight forward way to include the evolution of
the perturbations:
\begin{align}
C^{xy}_\ell(\chi,\chi') 
    \approx \frac{2}{\pi}
    \int_0^\infty\ud k k^2
    j_\ell(k\chi) j_{\ell}(k\chi')
    P_{xy}(k,(\chi + \chi')/2),\nonumber &\\
    |\chi - \chi'| \ll 1/aH.&
\end{align}

@@

Equation~\ref{e:Clss} can be simplified substantially by adopting the small
angle and Limber approximations \citep{1953ApJ...117..134L,
1992ApJ...388..272K, 1998ApJ...498...26K}.  The small angle
approximation eliminates the $k$ integral over spherical Bessel functions, replacing it
with a Fourier transform, and is valid for $\ell \gg 1$. The Limber
approximation assumes that only modes with small radial component of their wave
vector contribute to the radial integrals and is valid if the power spectra
evolve slowly compared to the correlation length \citep{2008PhRvD..78l3506L}
(which has already been
assumed). With these approximations we have
\begin{align}
    &C^{ss}_\ell(\chi, \chi') \approx
    \nonumber \\ & \quad
    \frac{1}{\bar\chi^2}
        \int_{-\infty}^\infty\frac{\ud k_\parallel}{(2 \pi)} 
        e^{i k_\parallel (\chi - \chi')}
        P_{[ee + ff - 2ef]}
        (\sqrt{k_\parallel^2 + \nu^2/\bar\chi^2}, \bar\chi)
    \nonumber \\ & \quad +
    A(\chi) A(\chi')
    \int_0^{\chi_{\rm min}}\ud\chi''
    \frac{1}{\chi^{\prime\prime 2}}
    P_{ee}(\nu/\chi'',\chi'')
    \nonumber \\ & \quad +
    \frac{A(\chi_{\max})}{\chi_{\min}^2}
    P_{[ee - ef]}(\nu/\chi_{\min}, \chi_{\min}),
    \label{e:Clss_limber}
\end{align}
where $\nu\equiv\ell + 1/2$, $\chi_{\min} \equiv \min(\chi, \chi')$,
and $\chi_{\max} \equiv \max(\chi,
\chi')$ .  We've found that these approximations are excellent for $\ell
\gtrsim 10$
and use this form for the remainder of the paper. In the above equation we dub
the three terms the `local', `integral' and `cross' terms respectively and will
refer to them as such for the remainder of this work.

@@

For plotting the DM space power-spectra we use a toy model for the electron and
FRB clustering that assumes both are biased tracers of the dark matter,
\emph{i.e.}~$P_{xy}(k)=b_x b_y P(k)$.  We
calculate the time dependant dark matter power-spectrum using {\tt CAMB} using
the integrated {\tt HALOFIT} to approximate the non-linear evolution. We assume
that the electron bias $b_e=1$, which should be true on large scales due to
approximate conservation of free electrons. The bias of FRB sources is
unconstrained by data and we choose it to be $b_f=1.3$. The coefficient
$A(\chi)$ depends on the density of detected FRB sources, $\bar n_f(\chi)$.
We use a simple model where all events above a fixed flux are detected,
the intrinsic luminosity function has the
\citet{1976ApJ...203..297S} form with index $-1$, and events at the cut off of the
luminosity function and at
a radial distance of $\chi=2350\,{\rm MPc}/h$ are at the detection threshold.
This source density is shown in Figure~\ref{f:n_f} with the resulting DM-space
power-spectrum in Figure~\ref{f:terms}.

\begin{figure}
    \includegraphics[scale=0.42]{figures/n_f}
    \caption{
        \label{f:n_f}
        \emph{Top:} Comoving density of sources in toy model, as described in
        text. Normalization is
        such that the survey has a total of $10^4$ FRBs.
        \emph{Bottom:} Resulting coefficient $A(\chi)$ as given in
        Equation~\ref{e:A}.
    }
\end{figure}

\begin{figure}
    \includegraphics[scale=0.42]{figures/terms}
    \caption{
        \label{f:terms}
        Terms in the DM-space cross-correlation angular power spectrum.
        All terms are evaluated at $\bar\chi=2000\,{\rm MPc}/h$ and 
        $\chi - \chi' = 10\,{\rm MPc}/h$, except the lowest most 
        ``local term'' curve which is for $\chi - \chi'=50\,{\rm MPc}/h$. 
        While
        the other two terms are highly insensitive to the separation of
        the radial slices, the local term drops rapidly with separation.
    }
\end{figure}

@@

I can be seen in Figure~\ref{f:terms} that the integral term dominates the
cross term and the local term except at small radial separations.  We've found
that this is generically true for most $(\chi, \chi')$ pairs and that the
sensitivity shown in Figure~\ref{f:sensitivity} is dominated by the integral
term. Since the integral term only depends on the clustering of electrons, not
sources, an FRB survey could be used to study the distribution of electrons and
potentially solve the `missing baryon problem' as pointed out in
\citet{2014ApJ...780L..33M}.

We have calculated the clustering of sources in dispersion measure space and
shown that a survey detecting a modest $10^4$ fast radio bursts of cosmological
origin could detect the clustering signal.

Looking forward, DM-space
clustering could become a powerful probe of large-scale structure should
observational factors turn out to be favourable compared to spectroscopic
surveys. The baryon acoustic oscillation feature can be clearly seen in
Figure~\ref{f:terms} in the cross term and both
local-term curves. It might even be possible to extract it from the integral
term using tomography.
This however depends on the contribution to dispersion measure from
the source's environment and host halo to be modest compared to the
contribution from the IGM.
